%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Acknowledgements %%%%%%%%%%%%%%%%%%%%

\begin{center}
\Large{\textbf{Acknowledgements}}
\end{center}

\vspace{\myparskip}

I would like to express my heartfelt gratitude for the support and
advice of Dr.~Paul~Callanan. Throughout my undergraduate degree,
Dr.~Callanan encouraged my interest in astrophysics and provided me
with the opportunity to gain experience in this field. His supervision
of my M.Sc.\ has helped cement my interest and has given me valuable
skills and knowledge. I am also indebted to him for his help in both
selecting and securing a university place for my Ph.D.\ studies. %

\vspace{\myparskip}

My thanks must also be given to the U.C.C.\ Physics Department. The
academic staff have stimulated my mind for several years, while the
technical and clerical staff have helped me with more practical
matters. I must not forget my fellow students, especially Lisa Goggin,
all of whom have helped me struggle through five years of Physics. %

\vspace{\myparskip}

Thanks also to the present and former occupants of the Astrophysics
Laboratory. Manuel Perez-Torres generously spent a lot of time
explaining X-ray Binaries to me, in spite of the demands of his own
Ph.D. John~McCarthy and Peter~Curran have both helped
me in my research work and in the preparation of this document. %

\vspace{\myparskip}

Finally, my sincere gratitude to my parents, Tom and Vera, and my
sister, Bridget, who have supported and encouraged me throughout my
entire education, and also to my dearest Gillian, who was very
patient and understanding. %
